---
title: "Diagonal Case - Multiple Stocks/Ages"
author: "Carl James Schwarz"
date: "`r Sys.Date()`"
output: 
  html_vignette: 
    toc: true # table of content true
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: true  ## if you want number sections at each table header
#vignette: >
#  %\VignetteIndexEntry{02 - Diagonal Case - Multiple Stocks/Ages}
#  %\VignetteEncoding{UTF-8}
#  %\VignetteEngine{knitr::rmarkdown_notangle}
editor_options: 
  chunk_output_type: inline
---

```{r setup, include = FALSE,message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(binom)
library(BTSPAS)
library(ggplot2)
max.width=70
```
# Location of vignette source and code.
Because of the length of time needed to run the vignettes, only
static vignettes have been included with this package.

The original of the vignettes and the code can be obtained from 
the GitHub site at
https://github.com/cschwarz-stat-sfu-ca/BTSPAS

# Introduction
In some cases, the population of fish consist of a mixture of ages (young of year, and juvenile) and
stocks (wild and hatchery). *BTSPAS* has a number of routines to handle three common
occurrences as explained below.

In all cases, only diagonal recoveries are allowed.

Refer to the vignette on the *Diagonal Case* for information about fixing values of $p$ or modelling 
$p$ using covariates such a stream flow or smoothing $p$ using a temporal spline.


# Wild and Hatchery Chinook

In each stratum $j$, $n1[j]$ fish are marked and released above a rotary screw trap.
Of these, $m2[j]$ are recaptured in the stratum of release, i.e. the matrix of
releases and recoveries is diagonal.
The $n1[j]$ and $m2[j]$ establish the capture efficiency of the trap.
 
At the same time, $u2[j]$ unmarked fish are captured at the screw trap.
These fish are a mixture of wild and hatchery raised chinook salmon.
A portion (*clip.rate*) of the hatchery raised fish are adipose fin clipped and can be recognized as hatchery raised.
The unclipped fish are a mixture of wild and hatchery fish which must be separated.
Hence the $u2[j]$ are separated into:

* $u2.A[j]$ representing the number of adipose clipped fish known to be hatchery fish, and
* $u2.N[j]$ representing the number of unclipped fish which are mixture of hatchery and wild fish.


## Reading in the data
Here is an example of some raw data that is read in:
```{r}
demo.data.csv <- textConnection(
"     jweek,      n1,      m2,      u2.A,      u2.N
          9,       0,       0,         0,      4135
         10,    1465,      51,         0,     10452
         11,    1106,     121,         0,      2199
         12,     229,      25,         0,       655
         13,      20,       0,         0,       308
         14,     177,      17,         0,       719
         15,     702,      74,         0,       973
         16,     633,      94,         0,       972
         17,    1370,      62,         0,      2386
         18,     283,      10,         0,       469
         19,     647,      32,         0,       897
         20,     276,      11,         0,       426
         21,     277,      13,         0,       407
         22,     333,      15,         0,       526
         23,    3981,     242,      9427,     30542
         24,    3988,      55,      4243,     13337
         25,    2889,     115,      1646,      6282
         26,    3119,     198,      1366,      5552
         27,    2478,      80,       619,      2959
         28,    1292,      71,       258,      1455
         29,    2326,     153,       637,      3575
         30,    2528,     156,       753,      4284
         31,    2338,     275,       412,      2903
         32,    1012,     101,       173,      1127
         33,     729,      66,        91,       898
         34,     333,      44,        38,       406
         35,     269,      33,        22,       317
         36,      77,       7,         8,       99
         37,      62,       9,         2,       77
         38,      26,       3,         4,        37
         39,      20,       1,         1,        22")

demo.data <- read.csv(demo.data.csv, header=TRUE, as.is=TRUE, strip.white=TRUE)

print(demo.data)
```

There are several unusual features of the data set: 

* No fish were tagged and released in the first stratum. So no information is available
to estimate the capture efficiency at the second trap in the first week.
* In one of the recovery strata, there were no recaptures and so the
estimated recapture probability will be zero. But the data shows that some unmarked
fish were captured in these strata, so the actual efficiency must have been non-zero.
* There one julian week  where the number of unmarked fish captured suddenly jumps by
several orders of magnitude. This
jump correspond to releases of hatchery fish into the system.
* Before the hatchery release, we know that all of the unmarked fish
without an adipose clip are wild fish.

## Fitting the  BTSPAS diagonal model

We already read in the data above. Here we set the rest of the parameters. 

* the *hatch.after* variable indicates the stratum after which hatchery fish are released
* the *bad.m2* variable identifies the stratum where the $m2$ value is unusual.
* the *clip.frac.H* variable identify what fraction of the hatchery fish have adipose fin clips.

Don't forget to set the working directory as appropriate:
```{r }
library("BTSPAS")  

# After which weeks do the hatchery fish start to arrive. Prior to this point, all fish are wild and it is not
# necessary to separate out the wild vs hatchery
demo.hatch.after <- c(22)  # julian weeks after which hatchery fish arrive.

# Which julian weeks have "bad"  values. These will be set to missing and estimated.
demo.bad.m2     <- c()   # list of julian weeks with bad m2 values
demo.bad.u2.A   <- c()     # list of julian weeks with bad u2.A values
demo.bad.u2.N   <- c()     # list of julian weeks with bad u2.N values

# The clipping fraction
demo.clip.frac.H <- .25    # what fraction of the hatchery fish are adipose fin clipped


# The prefix for the output files:
demo.prefix <- "JC-2003-CH"

# Title for the analysis
demo.title <- "Junction City 2003 Chinook - Separation of Wild and Hatchery YoY Chinook"


cat("*** Starting ",demo.title, "\n\n")

# Make the call to fit the model and generate the output files
demo.fit <- TimeStratPetersenDiagErrorWHChinook_fit(
                  title      =demo.title,
                  prefix     =demo.prefix,
                  time       =demo.data$jweek ,
                  n1         =demo.data$n1,
                  m2         =demo.data$m2,
                  u2.A       =demo.data$u2.A,
                  u2.N       =demo.data$u2.N ,
                  clip.frac.H= demo.clip.frac.H,
                  hatch.after=demo.hatch.after,
                  bad.m2     =demo.bad.m2, 
                  bad.u2.A   =demo.bad.u2.A, 
                  bad.u2.N   =demo.bad.u2.N,
                  debug=TRUE,  # this generates only 10,000 iterations of the MCMC chain for checking.
                  save.output.to.files=FALSE)

```

```{r echo=FALSE,results="hide" }
# delete extra files that were created
file.remove("data.txt"       )       
file.remove("CODAindex.txt"  )
file.remove("CODAchain1.txt" )
file.remove("CODAchain2.txt" )
file.remove("CODAchain3.txt" )
file.remove("inits1.txt"     )
file.remove("inits2.txt"     )
file.remove("inits3.txt"     )
file.remove("model.txt"      )

```

## The output from the fit

Here is the fitted spline curve to the number of unmarked fish available in each recovery sample
```{r message=FALSE,warning=FALSE,fig.width=6, fig.height=4, include=TRUE, fig.align="center", fig.cap=c("Fitted spline curve")}
demo.fit$plots$fit.plot
```

The separation of wild and hatchery fish is evident as well has the start of the spline for the hatchery fish.


A plot of the $logit(P)$ is 
```{r warnings=FALSE,message=FALSE,warning=FALSE, fig.width=6, fig.height=4,  include=TRUE, fig.align="center", fig.cap=c("Estimates of logit(p)")}
demo.fit$plots$logitP.plot
```
In cases where there is no information (such as the first julian week), $BTSPAS$ has interpolated based on the distribution of catchability
in the other strata and so the credible interval is very wide.

A summary of the posterior for each parameter is also available. In particular, here are the
summary statistics on the posterior sample for the total number unmarked separated by wild and hatchery origin:
```{r}
demo.fit$summary[ row.names(demo.fit$summary) %in% c("Ntot","Utot","Utot.H","Utot.W"),]
```



# Wild and Hatchery Chinook with YoY and Age 1 fish.

In this example, *BTSPAS* allows for
separating wild from hatchery Chinook salmon when Age-1 chinook are present (residualized) from last year.


In each stratum $j$, $n1[j]$ fish are marked and released above a rotary screw trap.
Of these, $m2[j]$ are recaptured in the stratum of release, i.e. the matrix of
releases and recoveries is diagonal.
The $n1[j]$ and $m2[j]$ establish the capture efficiency of the trap.
 
At the same time, $u2[j]$ unmarked fish are captured at the screw trap.
These fish are a mixture of YoY and Age-1 wild and hatchery raised chinook salmon. 
A portion (*clip.rate.H.YoY*, *clip.rate.H.1*) of the YoY and Age1 hatchery raised fish 
are adipose fin clipped and can be recognized as hatchery raised.
The unclipped fish are a mixture of wild and hatchery fish which must be separated.
Hence the $u2[j]$ are separated into 

* $u2.A.YoY[j]$ representing the number of YoY adipose clipped fish known to be hatchery fish, and
* $u2.N.YoY[j]$ representing the number YoY unclipped fish which are mixture of hatchery and wild fish
* $u2.A.1  [j]$ representing the number of Age1 adipose clipped fish known to be hatchery fish, and
* $u2.N.1  [j]$ representing the number of Age1 unclipped fish) which are mixture of hatchery and wild fish.


## Reading in the data
Here is an example of some raw data that is read in:
```{r}
demo2.data.csv <- textConnection(
"     jweek,      n1,      m2,      u2.A.YoY,      u2.N.YoY,      u2.A.1,      u2.N.1
          2,       0,       0,             0,            15,           0,          10
          3,       0,       0,             0,            94,           1,          90
          4,     833,      52,             0,           385,           2,         112
          5,     852,      67,             0,          1162,          12,         140
          6,    1495,      77,             0,           592,          10,         103
          7,    1356,     182,             0,          1151,           4,          94
          8,    1889,     145,             0,          2258,           7,         121
          9,    2934,      89,             0,          1123,           2,          80
         10,    1546,      53,             0,          2277,           5,          57
         11,    4001,     232,             0,          2492,           4,          27
         12,    2955,     158,             0,          1579,          14,          88
         13,     529,      14,             0,          1046,           5,          45
         14,    1172,      49,             0,           766,           3,          13
         15,    3204,     232,             0,          2702,           1,           9
         16,    1328,      57,             0,         10408,           2,          18
         17,    3540,     114,             0,         12145,           3,          15
         18,    4791,      45,             0,           186,           0,           1
         19,    4808,      11,             0,           407,           0,           2
         20,    5952,      44,             0,           862,           0,           0
         21,    3852,      55,             0,           465,           0,           0
         22,    2621,      17,             0,           724,           0,          27
         23,    2131,      37,           854,          4860,           0,           0
         24,    5002,     152,           794,          3539,           0,           1
         25,    3706,     120,           904,          4597,           0,           0
         26,    1225,      44,           708,          3819,           0,           0
         27,     723,      45,           762,          3300,           0,           0
         28,    2895,     167,          1356,          5460,           0,           0
         29,    1395,     117,           614,          2918,           0,           0
         30,     479,      77,           420,          2252,           0,           0
         31,     964,      74,           289,          1240,           0,           0
         32,    2803,     288,            87,           428,           0,           0
         33,     952,      51,           114,           464,           0,           0
         34,     880,     126,            53,           515,           0,           0
         35,       0,       0,            21,            93,           0,           0")

demo2.data <- read.csv(demo2.data.csv, header=TRUE, as.is=TRUE, strip.white=TRUE)

print(demo2.data)
```

There are several unusual features of the data set: 

* No fish were tagged and released in the first two strata nor in the last stratum. So no information is available
to estimate the capture efficiency at the second trap in these strata.
* There one julian week  where the number of unmarked fish captured suddenly jumps by
several orders of magnitude. This
jump correspond to releases of hatchery fish into the system.
* Before the hatchery release, we know that all of the unmarked fish
without an adipose clip are wild fish.

## Fitting the  BTSPAS diagonal model

We already read in the data above. Here we set the rest of the parameters. 

* the *hatch.after* variable indicates the stratum after which hatchery fish are released
* the *bad.m2* variable identifies the stratum where the $m2$ value is unusual.
* the *clip.frac.H* variables identify what fraction of the hatchery fish have adipose fin clips in the YoY and Age 1 fish.

Don't forget to set the working directory as appropriate:
```{r }
library("BTSPAS")  

# After which weeks do the YoY hatchery fish start to arrive. 
# Prior to this point, all YoY fish are wild and it is not
# necessary to separate out the YoY wild vs hatchery
demo2.hatch.after.YoY <- c(22)  # julian weeks after which YoY hatchery fish arrive.

# Which julian weeks have "bad"  values. These will be set to missing and estimated.
demo2.bad.m2       <- c()     # list of julian weeks with bad m2 values
demo2.bad.u2.A.YoY <- c()     # list of julian weeks with bad u2.A.YoY values
demo2.bad.u2.N.YoY <- c()     # list of julian weeks with bad u2.N.YoY values
demo2.bad.u2.A.1   <- c()     # list of julian weeks with bad u2.A.YoY values
demo2.bad.u2.N.1   <- c()     # list of julian weeks with bad u2.N.YoY values

# The clipping fraction for the current YoY and last year's YoY (which are now Age 1 fish)
demo2.clip.frac.H.YoY <- .25    # what fraction of the YoY  hatchery fish are adipose fin clipped
demo2.clip.frac.H.1   <- .25    # what fraction of the Age1 hatchery fish are adipose fin clipped

# The prefix for the output files:
demo2.prefix <- "NF-2009-CH-WH-YoY-Age1" 

# Title for the analysis
demo2.title <- "North Fork 2009 Chinook - Separation of YoY and Age 1 Wild and Hatchery Chinook"


cat("*** Starting ",demo2.title, "\n\n")

# Make the call to fit the model and generate the output files
demo2.fit  <- TimeStratPetersenDiagErrorWHChinook2_fit(
                  title    = demo2.title,
                  prefix   = demo2.prefix,
                  time     = demo2.data$jweek,
                  n1       = demo2.data$n1,          
                  m2       = demo2.data$m2,         
                  u2.A.YoY = demo2.data$u2.A.YoY,
                  u2.N.YoY = demo2.data$u2.N.YoY,
                  u2.A.1   = demo2.data$u2.A.1,
                  u2.N.1   = demo2.data$u2.N.1,
                  clip.frac.H.YoY= demo2.clip.frac.H.YoY,
                  clip.frac.H.1  = demo2.clip.frac.H.1,
                  hatch.after.YoY= demo2.hatch.after.YoY,
                  bad.m2       = demo2.bad.m2, 
                  bad.u2.A.YoY = demo2.bad.u2.A.YoY, 
                  bad.u2.N.YoY = demo2.bad.u2.N.YoY,
                  bad.u2.A.1   = demo2.bad.u2.A.1  , 
                  bad.u2.N.1   = demo2.bad.u2.N.1,
                  debug=TRUE,  # this generates only 10,000 iterations of the MCMC chain for checking.
                  save.output.to.files=FALSE)

```

```{r echo=FALSE,results="hide" }
# delete extra files that were created
file.remove("data.txt"       )       
file.remove("CODAindex.txt"  )
file.remove("CODAchain1.txt" )
file.remove("CODAchain2.txt" )
file.remove("CODAchain3.txt" )
file.remove("inits1.txt"     )
file.remove("inits2.txt"     )
file.remove("inits3.txt"     )
file.remove("model.txt"      )

```

## The output from the fit

Here is the fitted spline curve to the number of unmarked fish available of both stocks and ages.
```{r message=FALSE,warning=FALSE,fig.width=6, fig.height=4, include=TRUE, fig.align="center", fig.cap=c("Fitted spline curve")}
demo2.fit$plots$fit.plot
```

The separation of wild and hatchery fish is evident as well has the start of the spline for the hatchery fish.


A plot of the $logit(P)$ is 
```{r warnings=FALSE,message=FALSE,warning=FALSE, fig.width=6, fig.height=4,  include=TRUE, fig.align="center", fig.cap=c("Estimates of logit(p)")}
demo2.fit$plots$logitP.plot
```
In cases where there is no information (such as the first julian week), $BTSPAS$ has interpolated based on the distribution of catchability
in the other strata and so the credible interval is very wide.

A summary of the posterior for each parameter is also available. In particular, here are the
summary statistics on the posterior sample for the total number unmarked separated by wild and hatchery origin and the different ages:
```{r}
round(demo2.fit$summary[ grepl("Utot", row.names(demo2.fit$summary)),],1)
```



# Wild and Hatchery Steelhead with YoY and Age 1 fish.

In this analysis we fit a diagonal time-stratified Petersen estimator 
separating wind from hatchery Steelhead salmon..

This differs from the Wild vs Hatchery Chinook salmon in previous sections in that all hatchery raised steelhead are marked,
so there is complete separation by age and (wild/hatchery).
There are 3 population of interest, Wild.YoY, Hatchery.Age1+, and  Wild.Age1+.

This analysis is based on the analysis of California Junction City 2003 Steelhead data and is the example used
in the Trinity River Project.


In each stratum $j$, $n1[j]$ fish are marked and released above a rotary screw trap.
Of these, $m2[j]$ are recaptured in the stratum of release, i.e. the matrix of
releases and recoveries is diagonal.
The $n1[j]$ and $m2[j]$ establish the capture efficiency of the trap.
 
At the same time, $u2[j]$ unmarked fish are captured at the screw trap.
These fish are a mixture of wild and hatchery raised steelhead salmon. 
The $u2[j]$ are separated into 

* $u2.W.YoY[j]$ representing wild, YoY steelhead,
* $u2.W.1  [j]$ representing wild, age 1+ steelhead, and
* $u2.H.1  [j]$ representing hatchery, age 1+ steelhead.


## Reading in the data
Here is an example of some raw data that is read in:
```{r}
demo3.data.csv <- textConnection(
"     jweek,      n1,      m2,      u2.W.YoY,      u2.W.1,      u2.H.1
          9,       0,       0,             0,          58,           0
         10,       0,       0,             0,         357,           2
         11,       0,       0,             0,         720,           0
         12,     999,       5,             0,         850,        4643
         13,    1707,      13,            11,         585,        5758
         14,    1947,      39,             0,         532,        4220
         15,    2109,       7,             0,         873,        2328
         16,     972,       1,             0,         303,        1474
         17,     687,       0,             1,         291,         875
         18,       0,       0,            33,          12,          39
         19,       0,       0,            31,         101,          15
         20,       0,       0,            11,          47,          13
         21,       0,       0,            78,          49,          26
         22,       0,       0,            46,          44,          22
         23,       0,       0,            35,          50,          59
         24,       0,       0,            30,          38,          15
         25,       0,       0,           309,          58,           8
         26,       3,       0,           278,          36,           4
         27,       0,       0,           207,          13,           2
         28,       0,       0,           196,           5,           0
         29,       0 ,      0,           613,          12,           0
         30,       0,       0,           764,          15,           0
         31,       0,       0,           556,          11,           0
         32,       0,       0,           250,          12,           0
         33,       0,       0,           106,          13,           0
         34,       0,       0,           413,          12,           0
         35,       0,       0,           995,          28,           1
         36,       0,       0,           357,          10 ,          0
         37,       0,       0,           181,           8,          27
         38,       0,       0,            53,           3,           2
         39,       0,       0,            29,           2,           0
         40,       0,       0,             3,           0,           0
         41,       0,       0,             5,           0,           0
         42,       0,       0,            14,           4,           0
         43,       0,       0,             8,          10,           0
         44,       0,       0,            19,           7,           0
         45,       0,       0,            46,           4,           0
         46,       0,       0,           229,           7,          0")

demo3.data <- read.csv(demo3.data.csv, header=TRUE, as.is=TRUE, strip.white=TRUE)

print(demo3.data)
```

There are several unusual features of the data set: 

* Marking and releasing of steelhead took place in only a few weeks! The hierarchical model
will extrapolate outside these weeks to estimate the capture rate. This is likely a very dangerous thing
to do.
* There one julian week  where the number of unmarked fish captured suddenly jumps by
several orders of magnitude. This
jump correspond to releases of hatchery fish into the system.
* Before the hatchery release, we know that all of the unmarked fish
without an adipose clip are wild fish.

## Fitting the  BTSPAS diagonal model

We already read in the data above. Here we set the rest of the parameters. 

* the *hatch.after* variable indicates the stratum after which hatchery fish are released
* the *bad.m2* variable identifies the stratum where the $m2$ value is unusual.
* the *clip.frac.H* variables identify what fraction of the hatchery fish have adipose fin clips in the YoY and Age 1 fish.

Don't forget to set the working directory as appropriate:
```{r }
library("BTSPAS")  

# After which weeks do the hatchery fish start to arrive. Prior to this point, all fish are wild and it is not
# necessary to separate out the wild vs hatchery
demo3.hatch.after <- c(11)  # julian weeks after which hatchery fish arrive.

# Which julian weeks have "bad"  values. These will be set to missing and estimated.
demo3.bad.m2       <- c()     # list of julian weeks with bad m2 values
demo3.bad.u2.W.YoY <- c()     # list of julian weeks with bad u2.W.YoY values
demo3.bad.u2.W.1   <- c()     # list of julian weeks with bad u2.W.1   values
demo3.bad.u2.H.1   <- c()     # list of julian weeks with bad u2.H.1   values



# The prefix for the output files:
demo3.prefix <- "demo-JC-2003-ST-TSPDE-WH" 

# Title for the analysis
demo3.title <- "Junction City 2003 Steelhead - Separation of Wild and Hatchery YoY and Age 1+ Steelhead"



cat("*** Starting ",demo3.title, "\n\n")

# Make the call to fit the model and generate the output files
demo3.fit <- TimeStratPetersenDiagErrorWHSteel_fit(
                  title     = demo3.title,
                  prefix    = demo3.prefix,
                  time      = demo3.data$jweek,
                  n1        = demo3.data$n1,
                  m2        = demo3.data$m2,
                  u2.W.YoY  = demo3.data$u2.W.YoY, 
                  u2.W.1    = demo3.data$u2.W.1, 
                  u2.H.1    = demo3.data$u2.H.1,
                  hatch.after=demo3.hatch.after,
                  bad.m2     = demo3.bad.m2,
                  bad.u2.W.YoY= demo3.bad.u2.W.YoY,
                  bad.u2.W.1  = demo3.bad.u2.W.1,
                  bad.u2.H.1  = demo3.bad.u2.H.1,
                  debug=TRUE,  # this generates only 10,000 iterations of the MCMC chain for checking.
                  save.output.to.files=FALSE)

```
The final parameter (*save.output.to.files*) can be set to automatically to save plots and reports in files with the appropriate prefix in the working directory.  

```{r echo=FALSE,results="hide" }
# delete extra files that were created
file.remove("data.txt"       )       
file.remove("CODAindex.txt"  )
file.remove("CODAchain1.txt" )
file.remove("CODAchain2.txt" )
file.remove("CODAchain3.txt" )
file.remove("inits1.txt"     )
file.remove("inits2.txt"     )
file.remove("inits3.txt"     )
file.remove("model.txt"      )

```

## The output from the fit

Here is the fitted spline curve to the number of unmarked fish available of both stocks and ages.
```{r message=FALSE,warning=FALSE,fig.width=6, fig.height=4, include=TRUE, fig.align="center", fig.cap=c("Fitted spline curve")}
demo3.fit$plots$fit.plot
```

The separation of wild and hatchery fish is evident as well has the start of the spline for the hatchery fish.


A plot of the $logit(P)$ is 
```{r warnings=FALSE,message=FALSE,warning=FALSE, fig.width=6, fig.height=4,  include=TRUE, fig.align="center", fig.cap=c("Estimates of logit(p)")}
demo3.fit$plots$logitP.plot
```
In cases where there is no information (such as the first julian week), $BTSPAS$ has interpolated based on the distribution of catchability
in the other strata and so the credible interval is very wide.

A summary of the posterior for each parameter is also available. In particular, here are the
summary statistics on the posterior sample for the total number unmarked separated by wild and hatchery origin and the different ages:
```{r}
demo3.fit$summary[ grepl("Utot", row.names(demo3.fit$summary)),]
```



# References

Bonner, S. J., & Schwarz, C. J. (2011). 
Smoothing population size estimates for Time-Stratified Mark–Recapture experiments Using Bayesian P-Splines. 
Biometrics, 67, 1498–1507. 
https://doi.org/10.1111/j.1541-0420.2011.01599.x

Schwarz, C. J., & Dempson, J. B. (1994).
Mark-recapture estimation of a salmon smolt population. 
Biometrics, 50, 98–108.


