---
title: "Non-Diagonal Case"
author: "Carl James Schwarz"
date: "`r Sys.Date()`"
output: 
  html_vignette: 
    toc: true # table of content true
    toc_depth: 3  # up to three depths of headings (specified by #, ## and ###)
    number_sections: true  ## if you want number sections at each table header
#vignette: >
#  %\VignetteIndexEntry{03 - Non Diagonal Case}
#  %\VignetteEncoding{UTF-8}
#  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
---

```{r setup, include = FALSE,message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(binom)
library(BTSPAS)
library(ggplot2)
max.width=70
```
# Location of vignette source and code.
Because of the length of time needed to run the vignettes, only
static vignettes have been included with this package.

The original of the vignettes and the code can be obtained from 
the GitHub site at
https://github.com/cschwarz-stat-sfu-ca/BTSPAS

# Introduction

This case represents a generalization of the diagonal case considered in a separate
vignette. Now, rather than assuming that all recaptures from a release take place in 
a single recovery stratum, recoveries could take place over multiple recovery strata.  

Again, consider an experiment to estimate the number of outgoing smolts on a small river. The 
run of smolts extends over several weeks. As smolts migrate, they are captured and marked
with individually numbered tags and released at the first capture location using, for example, a
fishwheel. The migration continues, and a second fishwheel takes a second sample several
kilometers down stream. At the second fishwheel, the captures consist of a mixture of marked
(from the first fishwheel) and unmarked fish.

The efficiency of the fishwheels varies over time in response to stream flow, run size passing
the wheel and other uncontrollable events. So it is unlikely that the capture probabilities are
equal over time at either location, i.e. are heterogeneous over time. 

We suppose that we can temporally stratify the data into, for example, weeks, where the
capture-probabilities are (mostly) homogeneous at each wheel in each week.

But now, we allow
tagged animals to be captured in several recovery strata. 
For example, 
suppose that in each julian week $j$, $n1[j]$ fish are marked and released above the rotary screw trap.
Of these, $m2[j,j]$ are recaptured in julian week $j$;
$m2[j,j+1]$ are recaptured in julian week $j+1$;
$m2[j,j+2]$ are recaptured in julian week $j+2$ and so on.

At the same time, $u2[j]$ unmarked fish are captured at the screw trap.

This implies that the data can be structured
as a **non-diagonal** array similar to:

```{}
Recovery Stratum
               tagged    rs1      rs2     rs3 ...rs4                 rsk  rs(k+1)
Marking   ms1    n1[1]  m2[1,1] m2[1,2] m2[1,3] m2[1,4]      0  ...   0      0 
Stratum   ms2    n1[2]   0      m2[2,2] m2[2,3] m2[2,4] .... 0  ...   0      0 
          ms3    n1[3]   0       0      m2[3,3] m2[3,4] ...  0  ...   0      0  
          ...  
          msk    n1[k]   0       0      0  ...  0            0    m2[k,k] m2[k,k+1]  
Newly  
Untagged               u2[1]   u2[2]   u2[3]  ...                 u2[k]   u2[k,k+1]
captured  
```
Here the tagging and recapture events have been stratified in to $k$ temporal strata.
Marked fish from one stratum tend to spread out and are recaptured over multiple strata.
Several additional recovery strata are needed at the end of the experiment to fully
capture the final release stratum.

Because the lower diagonal of the recovery matrix is zero, the data can be entered
in a shorthand fashion by showing the recoveries in the same stratum as release,
the next stratum, etc, up to a maximum number of recovery strata per release.


# Example of basic non-diagonal BTSPAS fit.
## Reading in the data
Here is an example of some raw data that is read in:
```{r}
demo.data.csv <- textConnection(
"Date        ,     n1  , X0  , X1  , X2  , X3  , X4
 1987-04-26  ,      8  ,  0  ,  0  ,  0  ,  0  ,  2  
 1987-04-27  ,      5  ,  0  ,  0  ,  0  ,  0  ,  0  
 1987-04-28  ,      6  ,  0  ,  0  ,  0  ,  0  ,  0  
 1987-04-29  ,     17  ,  0  ,  0  ,  2  ,  1  ,  1 
 1987-04-30  ,     66  ,  0  ,  1  ,  0  ,  2  ,  3  
 1987-05-01  ,    193  ,  0  ,  1  ,  7  ,  7  ,  2  
 1987-05-02  ,     90  ,  0  ,  2  ,  0  ,  0  ,  0  
 1987-05-03  ,    260  ,  0  ,  0  , 14  ,  6  ,  1  
 1987-05-04  ,    368  ,  0  ,  9  , 46  ,  4  ,  2  
 1987-05-05  ,    506  ,  0  , 38  , 33  , 11  ,  0  
 1987-05-06  ,    317  ,  1  , 27  , 26  ,  3  ,  1  
 1987-05-07  ,     43  ,  0  ,  4  ,  3  ,  0  ,  2  
 1987-05-08  ,    259  ,  1  , 42  ,  5  ,  2  ,  0  
 1987-05-09  ,    259  ,  1  , 32  , 27  ,  1  ,  0  
 1987-05-10  ,    249  ,  1  , 85  ,  3  ,  1  ,  0  
 1987-05-11  ,    250  ,  3  , 21  , 19  ,  2  ,  0  
 1987-05-12  ,    298  , 42  , 16  , 11  ,  9  ,  1  
 1987-05-13  ,    250  ,  1  ,  7 ,  25  ,  6  ,  4  
 1987-05-14  ,    193  ,  0  ,  9 ,  18  ,  8  ,  0  
 1987-05-15  ,    207  ,  0  , 17  , 21  ,  2  ,  0  
 1987-05-16  ,    175  ,  0  , 18  , 10  ,  1  ,  0  
 1987-05-17  ,    141  ,  0  , 12  , 14  ,  7  ,  1  
 1987-05-18  ,    155  ,  0  ,  1  , 19  , 13  ,  6  
 1987-05-19  ,    123  ,  0  ,  5  , 22  ,  5  ,  0  
 1987-05-20  ,    128  ,  0  ,  6  , 17  ,  2  ,  1 
 1987-05-21  ,     72  ,  0  , 11  ,  9  ,  2  ,  0  
 1987-05-22  ,     57  ,  0  ,  6  ,  8  ,  0  ,  1  
 1987-05-23  ,     49  ,  0  ,  4  ,  2  ,  1  ,  0  
 1987-05-24  ,     57,   14  ,  2  ,  1  ,  0  ,  0  
 1987-05-25  ,     18  ,  0  ,  3  ,  0  ,  0  ,  0  
 1987-05-26  ,     20  ,  0  ,  3  ,  4  ,  0  ,  0  
 1987-05-27  ,     16  ,  0  ,  3  ,  0  ,  0  ,  0  
 1987-05-28  ,     15  ,  0  ,  0  ,  2  ,  0  ,  0  
 1987-05-29  ,     10  ,  0  ,  1  ,  0  ,  1  ,  0  
 1987-05-30  ,     13  ,  0  ,  0  ,  2  ,  0  ,  0  
 1987-05-31  ,      8  ,  0  ,  3  ,  1  ,  0  ,  0  
 1987-06-01  ,      2  ,  0  ,  1  ,  0  ,  0  ,  0  
 1987-06-02  ,     23  ,  0  ,  6  ,  0  ,  0  ,  0  
 1987-06-03  ,     20  ,  0  ,  2  ,  0  ,  0  ,  0  
 1987-06-04  ,     10  ,  0  ,  4  ,  1  ,  0  ,  0  
 1987-06-05  ,     10  ,  3  ,  1  ,  0  ,  0  ,  0 
 1987-06-06  ,      5  ,  0  ,  2  ,  0  ,  0  ,  1  
 1987-06-07  ,      2  ,  0  ,  0  ,  0  ,  0  ,  0  
 1987-06-08  ,      2  ,  0  ,  1  ,  0  ,  0  ,  0  ")

demo.data <- read.csv(demo.data.csv, header=TRUE, as.is=TRUE, strip.white=TRUE)

print(demo.data)
demo.data$Date <- as.Date(demo.data$Date, "%Y-%m-%d")
demo.data$jday <- as.numeric(format(demo.data$Date, "%j"))
```
Here the strata are days (rather than weeks). 
There are `r nrow(demo.data)` release strata, and tagged fish are recovered in the stratum of release plus another 
`r ncol(demo.data)-2` strata.

In the first release stratum, a total of `r demo.data[1,"n1"]` fish were tagged and released.
No recoveries occurred until 4 days later.

Because the recoveries take place in more strata than releases, the $u2$ vector is read in separately. 
Note that is must be sufficiently long
to account for the number of releases plus potential movement, for a length of 
`r nrow(demo.data)+ ncol(demo.data)-4`:
```{r }
demo.data.u2 <- c(0  ,    2  ,    1  ,  2 ,   39 , 226  , 75 , 129 , 120 , 380 ,
                  921, 1005  , 1181 ,1087 , 1108 ,1685  ,671 ,1766 , 636 , 483 , 
                  170,  269  ,  212 , 260 ,  154 , 145  , 99 ,  58 ,  74 ,  40 ,  
                   50,   59  ,   40 ,   9 ,   14 ,  13  , 22 ,  24 ,  33 ,  19 ,
                   12,    7  ,    4 ,   0 ,    0 ,  59  ,  0 ,   0 )
```

We also separate out the recoveries $m2$ into a matrix
```{r }
demo.data.m2 <- as.matrix(demo.data[,c("X0","X1","X2","X3","X4")])
```

```{r pp,message=FALSE,echo=FALSE}
Nhat <-BTSPAS::SimplePetersen( sum(demo.data$n1), sum(demo.data.m2), sum(demo.data.u2))
```
## Preliminary screening of the data

A pooled-Petersen estimator would add all of the marked, recaptured and unmarked fish
to give an estimate of `r formatC( round(Nhat$N.est), digits=0, width=20, format="f",big.mark=",")`.

Let us look at the data in more detail:

* In many of the release strata, there are small number of fish tagged and released.
* Most the recoveries occur soon after release, but there is longer right tail.
* It is difficult to estimate capture efficiency directly because of the smearing of 
releases over multiple recovery strata. 


Let us look at the pattern of unmarked fish captured at the second trap:
```{r umarked, echo=FALSE, fig.width=6, fig.height=4, include=TRUE, fig.align="center", fig.cap=c("Observed number of unmarked recaptures"), warning=FALSE}

plotdata <- data.frame(jday=min(demo.data$jday):(min(demo.data$jday)+length(demo.data.u2)-1),
                       u2  = demo.data.u2)

ggplot(data=plotdata, aes(x=jday, y=log(u2)))+
  ggtitle("Number of unmarked fish captured by julian day")+
  geom_point()+
  geom_line()+
  xlab("Julian day")+
  ylab("log(Number of unmarked fish captured (u2))")

```

There appears to be a gradual rise and fall in the number of unmarked fish with no sudden jumps. 
Something funny appears to have happened around julian day 160 -- I suspect that some pooling of data has occurred here.

*BTSPAS* provides two fitting routine:

* Non-diagonal but the movement of fish after tagging follows a log-normal distribution following release as explained in
Dempson and Schwarz (2009) and extended by Bonner and Schwarz (2011).
* Non-diagonal but a non-parametric movement distribution is assumes as explained in
Bonner and Schwarz (2011).


# Fitting the BTSPAS non-diagonal model with a log-normal movement distribution. 

Bonner and Schwarz (2011) developed a model with the following features.

* Log-normal distribution of the distribution of times between release and availability at the second
trap.
* A spline is used to smooth the total number of unmarked fish presenting themselves at the second trap
over the strata
* A hierarchical model for the capture-probabilities is assumed where individual stratum capture
probabilities are assumed to vary around a common mean.

The model also allows the user to use covariates to explain some of the variation in the 
capture probabilities in much the same way as the diagonal case.

The $BTSPAS$ package also has additional features and options:

* if $u2$ is missing for any stratum, the program will use the spline to interpolate for the
missing stratum.
* if $n1$ and/or the entire corresponding row of $m2$ are missing, the program will use the hierarchical model to interpolate the capture
probabilities for the missing strata.
* the program allows you specify break points in the underlying spline to account
for external events.
* sometimes bad thing happen.  The vector $bad.m2$ indicates which julian weeks something went wrong. 
In the above example, the
number of recoveries in julian week 41 is far below expectations and leads to impossible
Petersen estimate for julian week 41. 
Similarly, the vector $bad.u2$ indicates which julian weeks, the number of unmarked fish is suspect.
In both cases, the suspect values of $m2$ and $u2$ are set to missing.
Alternatively, the user can set the $m2$ and $u2$ values to missing in the data input directly.
I arbitrarily chose the third julian week to demonstrate this feature.

The $BTSPAS$ function also allows you specify

* The prefix is used to identify the output files for this run.
* The title  is used to title the output.
* Various parameters to control the Bayesian MCMC phase of model fitting. Please contact us for help in setting these
if problem arise.

We already read in the data above. Here we set the rest of the parameters. Don't forget to set the working directory as appropriate
```{r }
library("BTSPAS")  

# After which weeks is the spline allowed to jump?
demo.jump.after <- c()  # julian weeks after which jump occurs

# Which julian weeks have "bad" recapture values. These will be set to missing and estimated.
demo.bad.m2     <- c(117)   # list julian weeks with bad m2 values. 
demo.bad.u2     <- c()   # list julian weeks with bad u2 values. 
demo.bad.n1     <- c(117)   # list julian weeks with bad n1 values. 

# The prefix for the output files:
demo.prefix <- "demo-1987-Conne River-TSP NDE" 

# Title for the analysis
demo.title <- "Conne River 1987 Atlantic Salmon Smolts - Log-normal"

cat("*** Starting ",demo.title, "\n\n")

# Make the call to fit the model and generate the output files
demo.fit <- TimeStratPetersenNonDiagError_fit(
                  title=      demo.title,
                  prefix=     demo.prefix,
                  time=       min(demo.data$jday):(min(demo.data$jday)+length(demo.data.u2)-1),
                  n1=         demo.data$n1, 
                  m2=         demo.data.m2, 
                  u2=         demo.data.u2,
                  jump.after= demo.jump.after,
                  bad.n1=     demo.bad.n1,
                  bad.m2=     demo.bad.m2,
                  bad.u2=     demo.bad.u2,
                  debug=TRUE,             # save time by reducing number of MCMC iterations
                  save.output.to.files=FALSE)
```
The final parameter (*save.output.to.files*) can be set to automatically to save plots and reports in files with the appropriate prefix in the working directory.  

```{r deletefiles,echo=FALSE,results="hide" }
# delete extra files that were created
file.remove("data.txt"       )       
file.remove("CODAindex.txt"  )
file.remove("CODAchain1.txt" )
file.remove("CODAchain2.txt" )
file.remove("CODAchain3.txt" )
file.remove("inits1.txt"     )
file.remove("inits2.txt"     )
file.remove("inits3.txt"     )
file.remove("model.txt"      )

```


## The output from the fit
The final object has many components
```{r components, results="hide"}
names(demo.fit)
```
```{r echo=FALSE}
save.options <- options()
options(width=max.width)
names(demo.fit)
options(save.options)
```

The *plots* sub-object contains many plots:
```{r results='hide'}
names(demo.fit$plots)
```
```{r echo=FALSE}
save.options <- options()
options(width=max.width)
names(demo.fit$plots)
options(save.options)
```

In particular, it contains plots of the initial spline fit (*init.plot*), 
the final fitted spline (*fit.plot*), 
the estimated capture probabilities (on the logit scale) (*logitP.plot*),
plots of the distribution of the posterior sample
for the total unmarked and marked fish (*post.UNtot.plot*) 
and model diagnostic plots (goodness of fit (*gof.plot*), trace (*trace...plot*), and autocorrelation plots (*act.Utot.plot*).

These plots are all created using the $ggplot2$ packages, so the user can modify the plot (e.g. change titles etc).

The $BTSPAS$ program also creates a report, which includes information about the data used in the fitting,
the pooled- and stratified-Petersen estimates, a test for pooling, and summaries of the posterior. Only the first few lines
are shown below:
```{r }
head(demo.fit$report)
```

Here is the fitted spline curve to the number of unmarked fish available in each recovery sample
```{r message=FALSE,warning=FALSE,fig.width=6, fig.height=4, include=TRUE, fig.align="center", fig.cap=c("Fitted spline curve")}
demo.fit$plots$fit.plot
```


The distribution of the posterior sample for the total number unmarked and total abundance is available:
```{r fig.width=6, fig.height=4, include=TRUE, fig.align="center", fig.cap=c("Distribution of posterior samples")}
demo.fit$plots$post.UNtot.plot
```


A plot of the $logit(P)$ is 
```{r warnings=FALSE,message=FALSE,warning=FALSE, fig.width=6, fig.height=4,  include=TRUE, fig.align="center", fig.cap=c("Estimates of logit(p)")}
demo.fit$plots$logitP.plot
```
In cases where there is little information, $BTSPAS$ has shared information based on the distribution of catchability
in the other strata.

A summary of the posterior for each parameter is also available. In particular, here are the
summary statistics on the posterior sample for the total number unmarked and total abundance:
```{r}
demo.fit$summary[ row.names(demo.fit$summary) %in% c("Ntot","Utot"),]
```
This also includes the Rubin-Brooks-Gelman statistic ($Rhat$) on mixing of the chains and the effective sample size 
of the posterior (after
accounting for autocorrelation).

The estimated total abundance from $BTSPAS$ is 
`r formatC(round(demo.fit$summary[ "Ntot","mean"]), big.mark=",", digits=0, format="f")` (SD 
`r formatC(round(demo.fit$summary[ "Ntot","sd"  ]), big.mark=",", digits=0, format="f")` ) fish.

The model has a "base" log-normal distribution for the travel times between the release and recovery 
strata. The summary of the posterior distribution of its parameters for the log(travel time) are:
```{r}
round(demo.fit$summary[ row.names(demo.fit$summary) %in% c("baseMu","baseSd"),],3)
```

Each release stratum is allowed to have a travel time distribution that differ from this base travel time distribution,
by allowing the individual release stratum parameters to be sampled from a distribution around the above vales.
Posterior samples represent the mean and standard deviation of the log(travel time) between
the release and recovery strata. Here are the results for the first 5 strata:
```{r}
round(demo.fit$summary[ grepl("muLogTT", row.names(demo.fit$summary)),][1:5,],3)
round(demo.fit$summary[ grepl("sdLogTT", row.names(demo.fit$summary)),][1:5,],3)
```

It is also possible to see the probability of moving from release stratum $i$ to recovery stratum $j$ by looking
at the $Theta[i,j]$ values. Here are the transition probabilities for the first release stratum:
```{r}
round(demo.fit$summary[ grepl("Theta[1,", row.names(demo.fit$summary),fixed=TRUE),][1:10,],3)
```

The probabilities should sum to 1 for each release group.

Samples from the posterior are also included in the *sims.matrix*, *sims.array* and *sims.list* elements
of the results object. 

It is always important to do model assessment before accepting the results from the model fit. 
Please contact me for details on how to interpret
the goodness of fit, trace, and autocorrelation plots. 



# Fitting the BTSPAS non-diagonal model with a non-parametric movement distribution. 

Bonner and Schwarz (2011) developed a model with the following features.

* Non-parametric distribution of the distribution of times between release and availability at the second
trap.
* A spline is used to smooth the total number of unmarked fish presenting themselves at the second trap
over the strata
* A hierarchical model for the capture-probabilities is assumed where individual stratum capture
probabilities are assumed to vary around a common mean.

The model also allows the user to use covariates to explain some of the variation in the 
capture probabilities in much the same way as the diagonal case.

The $BTSPAS$ package also has additional features and options:

* if $u2$ is missing for any stratum, the program will use the spline to interpolate for the
missing stratum.
* if $n1$ and/or the entire corresponding row of $m2$ are missing, the program will use the hierarchical model to interpolate the capture
probabilities for the missing strata.
* the program allows you specify break points in the underlying spline to account
for external events.
* sometimes bad thing happen.  The vector $bad.m2$ indicates which julian weeks something went wrong. In the above example, the
number of recoveries in julian week 41 is far below expectations and leads to impossible
Petersen estimate for julian week 41. Similarly, the vector $bad.u2$ indicates which julian weeks, the number of unmarked fish is suspect.
In both cases, the suspect values of $m2$ and $u2$ are set to missing.
Alternatively, the user can set the $m2$ and $u2$ values to missing in the data input directly.
I arbitrarily chose the third julian week to demonstrate this feature.

The $BTSPAS$ function also allows you specify

* The prefix is used to identify the output files for this run.
* The title  is used to title the output.
* Various parameters to control the Bayesian MCMC phase of model fitting. Please contact us for help in setting these
if problem arise.

We already read in the data above. Here we set the rest of the parameters. Don't forget to set the working directory as appropriate
```{r }
library("BTSPAS")  

# After which weeks is the spline allowed to jump?
demo2.jump.after <- c()  # julian weeks after which jump occurs

# Which julian weeks have "bad" recapture values. These will be set to missing and estimated.
demo2.bad.m2     <- c()   # list julian weeks with bad m2 values. This is used in the Trinity Example
demo2.bad.u2     <- c()   # list julian weeks with bad u2 values. [This was arbitrary to demostrate the feature.]
demo2.bad.n1     <- c()   # list julian weeks with bad n1 values. [This was arbitrary to demonstrate the feature.]

# The prefix for the output files:
demo2.prefix <- "demo2-1987-Conne River-TSP NDE NP" 

# Title for the analysis
demo2.title <- "Conne River 1987 Atlantic Salmon Smolts - Non-parametric"

cat("*** Starting ",demo2.title, "\n\n")

# Make the call to fit the model and generate the output files
demo2.fit <- TimeStratPetersenNonDiagErrorNP_fit(  # notice change in function name
                  title=      demo2.title,
                  prefix=     demo2.prefix,
                  time=       min(demo.data$jday):(min(demo.data$jday)+length(demo.data.u2)-1),
                  n1=         demo.data$n1, 
                  m2=         demo.data.m2, 
                  u2=         demo.data.u2,
                  jump.after= demo2.jump.after,
                  bad.n1=     demo2.bad.n1,
                  bad.m2=     demo2.bad.m2,
                  bad.u2=     demo2.bad.u2,
                  debug=TRUE,             # save time by reducing number of MCMC iterations
                  save.output.to.files=FALSE)
```
The final parameter (*save.output.to.files*) can be set to automatically to save plots and reports in files with the appropriate prefix in the working directory.  

```{r deletefiles2,echo=FALSE,results="hide" }
# delete extra files that were created
file.remove("data.txt"       )       
file.remove("CODAindex.txt"  )
file.remove("CODAchain1.txt" )
file.remove("CODAchain2.txt" )
file.remove("CODAchain3.txt" )
file.remove("inits1.txt"     )
file.remove("inits2.txt"     )
file.remove("inits3.txt"     )
file.remove("model.txt"      )

```


## The output from the fit

Here is the fitted spline curve to the number of unmarked fish available in each recovery sample
```{r message=FALSE,warning=FALSE,fig.width=6, fig.height=4, include=TRUE, fig.align="center", fig.cap=c("Fitted spline curve")}
demo2.fit$plots$fit.plot
```


The distribution of the posterior sample for the total number unmarked and total abundance is available:
```{r fig.width=6, fig.height=4, include=TRUE, fig.align="center", fig.cap=c("Distribution of posterior samples")}
demo2.fit$plots$post.UNtot.plot
```


A plot of the $logit(P)$ is 
```{r warnings=FALSE,message=FALSE,warning=FALSE, fig.width=6, fig.height=4,  include=TRUE, fig.align="center", fig.cap=c("Estimates of logit(p)")}
demo2.fit$plots$logitP.plot
```
In cases where there is little information, $BTSPAS$ has shared information based on the distribution of catchability
in the other strata.

A summary of the posterior for each parameter is also available. In particular, here are the
summary statistics on the posterior sample for the total number unmarked and total abundance:
```{r}
demo2.fit$summary[ row.names(demo2.fit$summary) %in% c("Ntot","Utot"),]
```
This also includes the Rubin-Brooks-Gelman statistic ($Rhat$) on mixing of the chains and the effective sample size 
of the posterior (after
accounting for autocorrelation).

The estimated total abundance from $BTSPAS$ is 
`r formatC(round(demo2.fit$summary[ "Ntot","mean"]), big.mark=",", digits=0, format="f")` (SD 
`r formatC(round(demo2.fit$summary[ "Ntot","sd"  ]), big.mark=",", digits=0, format="f")` ) fish.

The estimated distribution function is allowed by vary by release stratum around a common "mean" distribution.
```{r }
probs <- demo2.fit$summary[grepl("movep", row.names(demo2.fit$summary)),  ]
round(probs,3)
```
So we expect that about `r round(probs[1,"mean"]*100,0)`% of fish will migrate to the second trap in the day of release;
about `r round(probs[2,"mean"]*100,0)`% of fish will migrate to the second trap in the second day after release etc.

The movement for each release stratum varies around this base distribution.
It is also possible to see the probability of moving from release stratum $i$ to recovery stratum $j$ by looking
at the $Theta[i,j]$ values. Here are the transition probabilities for the first release stratum:
```{r}
round(demo2.fit$summary[ grepl("Theta[1,", row.names(demo2.fit$summary),fixed=TRUE),],3)
```

The probabilities should also sum to 1 for each release group.

It is always important to do model assessment before accepting the results from the model fit. 
Please contact me for details on how to interpret
the goodness of fit, trace, and autocorrelation plots. 


# Prior information on the movement probabilities.
It is possible to impose prior information on the movement probabilities in both cases. This would be useful
in cases with VERY sparse data!

In the non-parametric case, specify a vector that gives the relative weight of belief of movement. 
These are similar to a Dirchelet-type prior where the values representing belief in the distribution of travel times. 
For example, $prior.muTT=c(1,4,3,2)$ represents a system where the maximum travel time is 3 strata after release with $1/10=.1$ 
of the animals moving in the stratum of release $4/10=.4$ of the animals taking 1 stratum to move etc 
So if $prior.muTT=c(10,40,30,20)$, this represent the same movement pattern but a strong degree of belief because all of the numbers
are larger. AN intuitive explanation is that the $sum(prior.muTT)$ represents the number of animals observed to 
make this travel time distribution.

Here we will fit a fairly strong prior on the movement probabilities:

```{r }
demo3.prior.muTT=c(10,50,30,5,5)
```
where the probability of movement in the stratum of release and subsequent strata is
```{r}
round(demo3.prior.muTT/sum(demo3.prior.muTT),2)
```


We already read in the data above. Here we set the rest of the parameters. Don't forget to set the working directory as appropriate
```{r }
library("BTSPAS")  

# After which weeks is the spline allowed to jump?
demo3.jump.after <- c()  # julian weeks after which jump occurs

# Which julian weeks have "bad" recapture values. These will be set to missing and estimated.
demo3.bad.m2     <- c()   # list julian weeks with bad m2 values. This is used in the Trinity Example
demo3.bad.u2     <- c()   # list julian weeks with bad u2 values. [This was arbitrary to demonstrate the feature.]
demo3.bad.n1     <- c()   # list julian weeks with bad n1 values. [This was arbitrary to demonstrate the feature.]

# The prefix for the output files:
demo3.prefix <- "demo3-1987-Conne River-TSP NDE NP- prior" 

# Title for the analysis
demo3.title <- "Conne River 1987 Atlantic Salmon Smolts - Non-parametric - Strong Prior"

cat("*** Starting ",demo3.title, "\n\n")

# Make the call to fit the model and generate the output files
demo3.fit <- TimeStratPetersenNonDiagErrorNP_fit(  # notice change in function name
                  title=      demo3.title,
                  prefix=     demo3.prefix,
                  time=       min(demo.data$jday):(min(demo.data$jday)+length(demo.data.u2)-1),
                  n1=         demo.data$n1, 
                  m2=         demo.data.m2, 
                  u2=         demo.data.u2,
                  prior.muTT= demo3.prior.muTT,  # prior on moements
                  jump.after= demo3.jump.after,
                  bad.n1=     demo3.bad.n1,
                  bad.m2=     demo3.bad.m2,
                  bad.u2=     demo3.bad.u2,
                  debug=TRUE,             # save time by reducing number of MCMC iterations
                  save.output.to.files=FALSE)
```
The final parameter (*save.output.to.files*) can be set to automatically to save plots and reports in files with the appropriate prefix in the working directory.  

```{r deletefiles3,echo=FALSE,results="hide" }
# delete extra files that were created
file.remove("data.txt"       )       
file.remove("CODAindex.txt"  )
file.remove("CODAchain1.txt" )
file.remove("CODAchain2.txt" )
file.remove("CODAchain3.txt" )
file.remove("inits1.txt"     )
file.remove("inits2.txt"     )
file.remove("inits3.txt"     )
file.remove("model.txt"      )

```


## The output from the fit

Here is the fitted spline curve to the number of unmarked fish available in each recovery sample
```{r message=FALSE,warning=FALSE,fig.width=6, fig.height=4, include=TRUE, fig.align="center", fig.cap=c("Fitted spline curve")}
demo3.fit$plots$fit.plot
```


The distribution of the posterior sample for the total number unmarked and total abundance is available:
```{r fig.width=6, fig.height=4, include=TRUE, fig.align="center", fig.cap=c("Distribution of posterior samples")}
demo3.fit$plots$post.UNtot.plot
```


A plot of the $logit(P)$ is 
```{r warnings=FALSE,message=FALSE,warning=FALSE, fig.width=6, fig.height=4,  include=TRUE, fig.align="center", fig.cap=c("Estimates of logit(p)")}
demo3.fit$plots$logitP.plot
```
In cases where there is little information, $BTSPAS$ has shared information based on the distribution of catchability
in the other strata.

A summary of the posterior for each parameter is also available. In particular, here are the
summary statistics on the posterior sample for the total number unmarked and total abundance:
```{r}
demo3.fit$summary[ row.names(demo3.fit$summary) %in% c("Ntot","Utot"),]
```
This also includes the Rubin-Brooks-Gelman statistic ($Rhat$) on mixing of the chains and the effective sample size 
of the posterior (after
accounting for autocorrelation).

The estimated total abundance from $BTSPAS$ is 
`r formatC(round(demo3.fit$summary[ "Ntot","mean"]), big.mark=",", digits=0, format="f")` (SD 
`r formatC(round(demo3.fit$summary[ "Ntot","sd"  ]), big.mark=",", digits=0, format="f")` ) fish.

The estimated distribution function is allowed by vary by release stratum around a common "mean" distribution.
```{r }
probs <- demo3.fit$summary[grepl("movep", row.names(demo3.fit$summary)),  ]
round(probs,3)
```
So we expect that about `r round(probs[1,"mean"]*100,0)`% of fish will migrate to the second trap in the day of release;
about `r round(probs[2,"mean"]*100,0)`% of fish will migrate to the second trap in the second day after release etc.

The movement for each release stratum varies around this base distribution.
It is also possible to see the probability of moving from release stratum $i$ to recovery stratum $j$ by looking
at the $Theta[i,j]$ values. Here are the transition probabilities for the first release stratum:
```{r}
round(demo3.fit$summary[ grepl("Theta[1,", row.names(demo3.fit$summary),fixed=TRUE),],3)
```

The probabilities should also sum to 1 for each release group.

It is always important to do model assessment before accepting the results from the model fit. 
Please contact me for details on how to interpret
the goodness of fit, trace, and autocorrelation plots. 




# References

Bonner, S. J., & Schwarz, C. J. (2011). 
Smoothing population size estimates for Time-Stratified Mark–Recapture experiments Using Bayesian P-Splines. 
Biometrics, 67, 1498–1507. 
https://doi.org/10.1111/j.1541-0420.2011.01599.x

Schwarz, C. J., & Dempson, J. B. (1994).
Mark-recapture estimation of a salmon smolt population. 
Biometrics, 50, 98–108.


